\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-11-01 paper.tex
Various updates to intro

\begin{abstract}
New developments have made it possible to infer genetic genealogies in
the presence of recombination at scale, enabling many
downstream applications in population and statistical genetics.
The structure representing such recombinant genetic ancestry
is usually referred to as an ancestral recombination graph (ARG).
% Be specific about what we're talking about.
ARGs are classically defined by the graph of
% We can play a bit loose here, please don't change to "common ancestor"
coalescence and recombination events in the history of a sample of genomes,
with the differential inheritance structure encoded by the
crossover breakpoints associated with recombination events.
We show that this ``event ARG'' (eARG) encoding
is limited in the patterns of genetic inheritance that can be represented,
being insufficient to describe the outputs of modern ARG simulation
and inference methods,
and ill-suited to capturing the complex patterns of
relatedness that exist in contemporary datasets.
We propose an alternative ``genome ARG'' (gARG) encoding
expressed in terms of
% we haven't define a "genome" yet, so need to give a bit of context
specific genomes and their intervals of genetic inheritance.
We show that the gARG encoding can fully capture the richness of modern large-scale datasets
and is compatible with modern simulation and inference methods.
Moreover, we show that, unlike the eARG encoding which requires
% TODO rephrase "to be detailed" is ambiguous
all recombination events to be detailed,
the gARG encoding enables fine-grained
levels of precision about recombination to be represented,
opening up a rich set of additional details.
\end{abstract}

@@

\linenumbers
\section{Introduction}
\label{sec-intro}
% Para 1: ARGs are cool!
Estimating the genetic genealogy of a set of genome sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
Broadly speaking, an ARG describes the different paths of genetic inheritance
caused by recombination, and encodes a sequence of correlated genealogical
trees along the genome.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,
fan2022genealogical,
hejase2022deep,
guo2022recombination,
zhang2023biobank,
nowbandegani2023extremely,
ignatieva2023distribution,
fan2023likelihood}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.
\citet{lewanski2023era} provide a biologically oriented introduction to ARGs
and further discuss their potential impacts.

% Para 2: What are they, though?
Although it is widely accepted that ARGs are important, there is some
confusion about what, precisely, an ARG \emph{is}.
% The grammar is a nightmare here - the stochastic process is singular, right?
In its original form, the ARG is an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages was formalised as a
graph by Griffiths and colleagues~\citep{griffiths1991two,
ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
Subsequently, ARGs have come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
a \emph{realisation} of such a random process,
or an inference from a set of sampled genome sequences.
The distinction between stochastic process
and data structure is not clear cut, however,
and subfields use the term
differently (Appendix~\ref{sec-arg-history}).
Another concept of an ARG that we must be careful to distinguish
is that of the \emph{actual} ARG underlying a sample of genomes
and ultimately an entire species. This ARG is the true history,
perfectly resolved into binary splits and mergers by
the cellular processes of meiosis and mitosis, regardless
of sampling density or population processes
(Appendix~\ref{sec-cell-lineages-and-args}).
Just as this true, species ARG unquestionably exists,
it can also never be entirely known.
% TODO: I'm not happy with the phrasing here, anyone want to jump in?
Even if a species
had a sufficiently high mutation rate to identify every
recent branching point uniquely (and somehow managed to survive)
this high rate of mutations would inevitably obscure
more ancient structure.
% Must be a citation for this?
Similarly, it seems unlikely that the details
of ancient recombinations can ever be known with much precision.

% Popgen models and the species ARG
Species-level ARGs may seem fanciful, but the scale of
modern datasets makes it necessary for us to grapple with the idea.
The UK Biobank (UKB), for example, has
genotype data for around 500,000 humans~\citep{bycroft2018genome},
and exome~\citep{backman2021exome} and
whole genome sequence~\citep{halldorsson2022sequences} data
for large subsets of this cohort.
UKB is just one of many such projects~\citep[e.g.][]{turnbull2018hundred,
karczewski2020mutational,tanjo2021practical}, and
agricultural datasets are even more
detailed~\citep[e.g.][]{hayes20191000,Ros-Freixedes2020}.
While we are not sequencing the entire species yet,
we are certainly sampling at a scale where the assumptions of
population genetics models are substantially broken.
By necessity and design, such mathematical models are caricatures
of biological reality,
allowing us to draw general conclusions about population processes
without the burden of unnecessary detail.
The coalescent with recombination (henceforth, ``coalescent'')
is a powerful model that defines a distribution of ARGs
for a given set of parameters.
A key assumption of the coalescent is that
sample size is much less than the effective population size ($N_e$)
and although the model has proved surprisingly robust to violations
of this assumption~\citep{wakeley2012gene,bhaskar2014distortion,
nelson2020accounting},
sample size can now be an order of magnitude or more \emph{larger} than the
usually assumed $N_e$ values.
A model assuming a single $N_e$ would be a
drastic over-simplification in these cases of course, but
even if sufficiently complex demographic models~\citep{gower2022demes}
encompassing hundreds of populations, explosive growth rates and myriad
interconnections of migration, were somehow estimated and provided as input,
ARGs sampled from the coalescent cannot capture the complexities
of family structure or the vagaries of history inherent in these
datasets.

% What is the problem?
The true ARG underlying a modern population-scale dataset
therefore cannot be known in its entirety, and the value
of attempting to describe the intricate details of the recent past
through mathematical models is questionable.
A pragmatic goal that we may set, then, is to infer \emph{aspects}
of the true ARG and to capture uncertainty where it inevitably exists.
This is the approach taken by recent inference
% TODO should we mention SARGE? I don't think it's in the same class, but
% maybe I'm wrong?
methods~\citep{kelleher2019inferring,speidel2019method,zhang2023biobank},
which estimate an approximate version of the true
ARG, informed, but not strictly constrained, by population genetic
models and parsimony.
While these methods have made substantial progress there
are, however, some significant problems in this emerging field.
Our terminology and mathematical formalisms for ARGs,
ultimately derived from the coalescent,
are ill-suited to describing these approximate structures.
% TODO add some citations of examples?
Discussions are forced to contrast a ``full'' ARG, containing
a precise description of every recombination or coalescence event,
with a sequence of disconnected local trees.
This view is depauperate, obscuring many shades of important
and interesting detail between these two extremes.
Although the term ``ARG'' is now often used in a general
sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
informally
encompassing the varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient,
zhang2023biobank},
there is no corresponding mathematical description of an ARG data structure.
Classical formalisms, based on the Griffiths approach, are insufficient to
describe these rich details.
Given that ARGs have already been estimated for the 500,000
humans in UKB~\citep{kelleher2019inferring,zhang2023biobank} and
more than a million SARS-CoV-2 genomes~\citep{zhan2023towards},
it is vital that our formal methods and terminology catch up
with this reality.

We address these problems here by providing a simple formal description of an
ARG data structure, based on recording the intervals of
genetic inheritance between specific genomes.
We call this the ``genome ARG'', or gARG encoding. We
contrast this with the classical formal definition of an ARG, based on
recording common ancestor and recombination events, which we refer to as the
``event ARG'' or eARG encoding. We show that the new gARG encoding it is a
substantial generalisation of the classical eARG approach,  providing much more
flexibility in how genetic inheritance can be represented, and encompassing the
outputs of modern methods. We show that the gARG approach can represent many
different types of approximation, in particular allowing us to systematically
describe uncertainty about the temporal ordering of multiple recombinations. It
is important to note that throughout we are interested in the details of these
competing mathematical formulations and their practical consequences.
% Do we need to say this? Probably not?
We are not questioning the form of the true ARG, or its existance.

% A data structure is a means of encoding information, and
% a precise description of the structure allows us to mathematically
% reason about its properties.

@@

In many settings, nodes are dated, i.e.\ each
node $u\in N$ is associated with a time $\tau_u$.
The topological ordering defined by the directed graph structure
and an arrow of time (telling us which direction is pastwards)
is sufficient for many applications, however,
and we assume node dates are not known here.
% This should be obvious, but seems like it needs to be said
In practical settings, we will wish to associate additional
metadata with nodes such as sample identifiers or quality-control metrics.
It is therefore best to think of the
integers used here in the definition of a node as an \emph{identifier},
with which arbitrary additional information can be associated.
% Note: I don't see why we have to apologise for defining nodes as
% integers, that's just the definition. I don't think anyone really
% has a problem with this.
% Concretely defining nodes as integers in this way has some
% significant advantages. In particular, it provides a direct link
% between an ARG and its local trees in terms of a well-defined
% combinatorial object (Section~\ref{sec-ARG-and-local-trees}),
% as well as enabling straightforward connection between
% definitions and practical implementation (Section~\ref{sec-efficiency}).

@@

There is also a certain clarity gained by explicitly modelling nodes
in the inheritance graph as genomes.
Outside of the context of a
% The point being, events are perfectly well defined in the models
% but not in reality
mathematical model, an ``event'' is a slippery concept.
For example, \emph{which} genome along a chain of cell divisions should be
regarded as the one where an event occurred
(Appendix~\ref{sec-cell-lineages-and-args}),
or whether multiple recombinations
within a single individual should be regarded as one or multiple events are
debatable points.
% If we want to have an ARG software ecosystem then such fireside
% discussions won't help.
From the perspective of a concrete data structure,
ideally forming the basis of an ecosystem of interoperable
inference and analysis methods, such debates
are unproductive.

@@

Efficiently propagating and resolving ancestral material for
a sample through a pre-existing graph is a well-studied problem,
and central to recent advances in individual-based forward-time
simulations~\citep{kelleher2018efficient,haller2018tree}.
In contrast to the usual ``retrospective'' view of ARGs
discussed so far, these methods record an ARG forwards in
time in a ``prospective'' manner. Genetic inheritance relationships
are recorded exhaustively, generation-by-generation,
leading to a rapid build-up of information, much of which
will not be relevant to the genetic ancestry of the current population.
This redundancy is periodically removed using the ``simplify''
algorithm~\citep{kelleher2018efficient}, which propagates and
resolves ancestral material.
Simplification allows efficient subsetting of ARGs (removing both samples and
certain levels of detail about recombination, as discussed in
Section~\ref{sec-ARG-simplification}), and is also the key enabling factor when
incorporating prospective ARGs into
forward-time simulation.
% Define "sample-resolved" here for later use
We refer to a gARG that has been simplified with respect to a set of
samples, such that the inheritance annotations on its edges contain
no non-ancestral material,  as sample-resolved.

@@

\section{ARGs and local trees}
\label{sec-ARG-and-local-trees}
% (We use ``local'' rather than
% ``marginal''
% % could list loads here, but not much point.
% trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping,kelleher2016efficient}
% to avoid the potential confusion with the
% statistical concept of marginal distributions.)
The relationship between an ARG and its corresponding
local trees is subtle and important.
A fundamental property of genetics is that a
given DNA nucleotide is inherited from exactly one parent genome,
both at an organismal and cell-by-cell level
(Appendix~\ref{sec-cell-lineages-and-args}).
These paths of single-parent inheritance give rise,
by definition, to a tree structure.
As a result of recombination, adjacent nucleotides can have
different paths of inheritance, and
an ARG encodes the entire ensemble of local trees along the
% Said "sample of genomes" originally, but using term "genome" in two
% different ways in the same sentence then.
genome for a given set of sample nodes.
% This is for Yan - to me this is a given, but maybe some people feel
% the following is overly formal?
Precisely defining the process by which local trees are extracted
from an ARG is essential to our understanding of how ARGs and local trees
are related, and we require a concrete mathematical structure
to describe the local trees.

@@

Inferred ARGs are based on the \citet{kreitman1983nucleotide} dataset,
a standard benchmark in the classical ARG
literature.
It consists of 43 biallelic SNPs spanning 2.4Kb of the \emph{D.~melanogaster} ADH
locus on chromosome 2L. Where required for inference purposes we assume
mutation and recombination rates of $5.49\times10^{-9}$
and $2.40463\times10^{-9}$ per site per generation
\citep{schrider2013rates,comeron2012many}
and a constant effective population size of $1720600$ \citep{li2006inferring}, all
values being taken from the \texttt{stdpopsim} library
\citep{adrion2020community,lauterbur2023expanding}.
Software versions were \kwarg\ v1.0, \argweaverD\ (2019),
\tsinfer v0.3.1, and \relate v1.1.9: full details and replicable code for generating
these figures are available on GitHub (see Data Availability).

@@

Interest in ARG inference methods and downstream applications is
burgeoning, with exciting developments arriving at ever-increasing pace.
Without agreement on basic terminology and some standardisation
on data formats, however, the ARG revolution may falter.
For ARG-based methods to achieve mainstream status, we require
a rich supporting software ecosystem.
Ideally, this would comprise a wide range of different
inference methods specialised to different organisms, types
and scales of data and
% I'm thinking about things like Espalier specifically wanting
% to characterise the recombs, vs Relate mainly wanting to
% get good local trees
targeting particular inference goals, sharing a common ARG interchange format.
The output of these diverse inference methods could then be
processed by numerous downstream applications,
enabling users to freely compare the outputs of different
inference methods on their data, and the result of different
analyses on these inferences.
Earlier efforts to standardise ARG interchange shared this vision,
but did not succeed~\citep{cardona2008extended,mcgill2013graphml}.
Without a well-defined and efficient shared format
for ARG data interchange,
users will suffer from problems with converting data from one program's
output to another, an unfortunate hallmark of population genetics
software for many years~\citep{excoffier2006computer}.
Current methods tend to tightly couple downstream analysis
of the inferred ARG with the inference
itself within the same software package,
which is ultimately not compatible with the widespread use
of ARGs for routine data analysis, and a healthy and diverse software ecosystem.
The gARG encoding described here is a significant generalisation
of classical concepts, capable of describing even the bewildering complexity
of contemporary datasets and capturing important forms of temporal uncertainty.
The \texttt{tskit} library is a high-quality open-source gARG implementation,
with proven efficiency and
scalability~\citep[e.g.][]{anderson2022genes,zhan2023towards},
that is already in widespread use.
Adopting it as a community standard,
and utilising its rich suite of operations,
may accelerate progress and help to
establish ARG methods within mainstream genetics.

>>>>

\begin{abstract}
New developments have made it possible to infer genetic genealogies in
the presence of recombination at scale, enabling many
downstream applications in population and statistical genetics.
The structure representing such recombinant genetic ancestry
is usually referred to as an ancestral recombination graph (ARG).
% Be specific about what we're talking about.
An ARG is classically defined as a graph of
% We can play a bit loose here, please don't change to "common ancestor"
coalescence and recombination events in the history of a sample of genomes,
encoding an inheritance structure where
crossover breakpoints are associated with recombination events.
We show that this ``event ARG'' (eARG) encoding
is limited in the patterns of genetic inheritance that can be represented,
being insufficient to describe the outputs of recent ARG simulation
and inference methods,
and ill-suited to capturing the complex patterns of
relatedness that exist in contemporary datasets.
We propose an alternative ``genome ARG'' (gARG) encoding
expressed in terms of
% we haven't define a "genome" yet, so need to give a bit of context
specific genomes and their intervals of genetic inheritance.
We show that the gARG encoding can fully capture the richness of large-scale datasets
and is compatible with modern simulation and inference methods.
We also show that, unlike the eARG encoding which requires
% TODO rephrase "to be detailed" is ambiguous
all recombination events to be encoded,
the gARG encoding enables fine-grained
levels of precision about recombination to be represented,
opening up a rich set of additional details.
\end{abstract}

@@

\linenumbers
\section{Introduction}
\label{sec-intro}
% Para 1: ARGs are cool!
% NOTE: slight inaccuracy of "DNA" here offset by making the problem concrete.
Estimating the genetic genealogy of a set of DNA sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
Broadly speaking, an ARG describes the different paths of genetic inheritance
caused by recombination, and encodes a sequence of correlated genealogical
trees along the genome
(see \citet{lewanski2023era} for a biologically oriented introduction).
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,
fan2022genealogical,
hejase2022deep,
guo2022recombination,
zhang2023biobank,
nowbandegani2023extremely,
ignatieva2023distribution,
fan2023likelihood}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.

% Para 2: What are they, though?
Although it is widely accepted that ARGs are important, there is some
confusion about what, precisely, an ARG \emph{is}.
% The grammar is a nightmare here - the stochastic process is singular, right?
In its original form, the ARG is an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages was formalised as a
graph by Griffiths and colleagues~\citep{griffiths1991two,
ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
Subsequently, ARGs have come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
a \emph{realisation} of such a random process,
or an inferred ancestry of a sample of genomes.
The distinction between stochastic process
and data structure is not clear cut, however,
and subfields use the term
differently (Appendix~\ref{sec-arg-history}).
Another concept of an ARG that we must be careful to distinguish
is that of the \emph{actual} ARG underlying a sample of genomes
and ultimately an entire species or higher grouping. 
This ARG is the true history,
perfectly resolved into binary splits and mergers by
the cellular processes of meiosis and mitosis, regardless
of sampling density or population processes
(Appendix~\ref{sec-cell-lineages-and-args}).
Just as this true ARG unquestionably exists,
it can also never be entirely known.
% TODO: I'm not happy with the phrasing here, anyone want to jump in?
Even if the mutation rate were high enough to identify every
recent branching point uniquely (while remaining biologically survivable),
such a high rate would saturate the genome with mutations, inevitably obscuring
past history.
% Must be a citation for this?
Similarly, it is unlikely that the details
of ancient recombinations can ever be known with much precision.

% Popgen models and the species ARG
Species-level ARGs may seem fanciful, but the scale of
modern datasets makes it necessary for us to grapple with the idea.
The UK Biobank (UKB), for example, has
genotype data for around 500,000 humans~\citep{bycroft2018genome},
and exome~\citep{backman2021exome} and
whole genome sequence~\citep{halldorsson2022sequences} data
for large subsets of this cohort.
UKB is just one of many such projects~\citep[e.g.][]{turnbull2018hundred,
karczewski2020mutational,tanjo2021practical}, and
agricultural datasets are even
richer in terms of generations~\citep[e.g.][]{hayes20191000,Ros-Freixedes2020}.
While we are not sequencing entire species,
we are certainly sampling at a scale where the assumptions of
population genetics models are substantially broken.
By necessity and design, such mathematical models are caricatures
of biological reality,
allowing us to draw general conclusions about population processes
without the burden of unnecessary detail.
 % TODO we're singling out the coalescent here, but it's just an 
 % exmaple. The DTWF is no better an approximation of the last 5
 % generations of UKB, e.g.
The coalescent with recombination (henceforth, ``coalescent'')
is a powerful model that defines a distribution over ARGs
for a given set of parameters.
A key assumption of the coalescent is that
sample size is much less than the effective population size ($N_e$)
and although the model has proved surprisingly robust to violations
of this assumption~\citep{wakeley2012gene,bhaskar2014distortion,
nelson2020accounting},
sample size can now be an order of magnitude or more \emph{larger} than the
usually assumed $N_e$ values.
A model assuming a single $N_e$ would be a
drastic over-simplification in these cases of course, but
even if sufficiently complex demographic models~\citep{gower2022demes}
encompassing hundreds of populations, explosive growth rates and myriad
interconnections of migration, were somehow estimated and provided as input,
ARGs sampled from the coalescent cannot capture the full complexities
of family structure or the vagaries of history inherent in these
datasets.

% What is the problem?
The true ARG underlying a modern population-scale dataset
therefore cannot be known in its entirety, and the value
% TODO rephrase heavily!
of attempting to describe the intricate details of the recent past
through mathematical models is questionable.
A pragmatic goal that we may set, then, is to infer \emph{aspects}
of the true ARG and to capture uncertainty where it inevitably exists.
This is the approach taken by recent inference
% TODO should we mention SARGE? I don't think it's in the same class, but
% maybe I'm wrong?
methods~\citep{kelleher2019inferring,speidel2019method,zhang2023biobank},
which estimate an approximate version of the true
ARG, informed, but not strictly constrained, by population genetic
models and parsimony.
While these methods have made substantial progress there
are, however, some significant problems in this emerging field.
Our terminology and mathematical formalisms for ARGs,
ultimately derived from the coalescent,
are ill-suited to describing these approximate structures.
% TODO add some citations of examples?
Discussions are forced to contrast a ``full'' ARG, containing
a precise description of every recombination or coalescence event,
with a sequence of disconnected local trees.
This view is depauperate, obscuring many shades of important
and interesting detail between these two extremes.
Although the term ``ARG'' is now often used in a general
sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
informally
encompassing the varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient,
zhang2023biobank},
there is no corresponding mathematical description of an ARG data structure.
Classical formalisms, based on the Griffiths approach, are insufficient to
describe these rich details.
Given that ARGs have already been estimated for the 500,000
humans in UKB~\citep{kelleher2019inferring,zhang2023biobank} and
more than a million SARS-CoV-2 genomes~\citep{zhan2023towards},
it is vital that our formal methods and terminology catch up
with this reality.

We address these problems here by providing a simple formal description of an
ARG data structure, based on recording the intervals of
genetic inheritance between specific genomes.
We call this the ``genome ARG'', or gARG encoding. We
contrast this with the classical formal definition of an ARG, based on
recording common ancestor and recombination events, which we refer to as the
``event ARG'' or eARG encoding. We show that the new gARG encoding is a
substantial generalisation of the classical eARG approach,  providing much more
flexibility in how genetic inheritance can be represented, and encompassing the
outputs of modern methods. We show that the gARG approach can represent many
different types of approximation, in particular allowing us to systematically
describe uncertainty about the temporal ordering of multiple recombinations. It
is important to note that throughout we are interested in the details of these
competing mathematical formulations and their practical consequences.
% Do we need to say this? Probably not?
We are not questioning the form of the true ARG, or its existence.

@@

In many settings, nodes are dated, i.e.\ each
node $u\in N$ is associated with a time $\tau_u$. 
A node can represent any genome along a 
chain of cell divisions~\ref{sec-cell-lineages-and-args},
or can be interpreted as representing one of ``the'' genomes of an
individual. How we assign dates will therefore vary according to application.
The topological ordering defined by the directed graph structure
and an arrow of time (telling us which direction is pastwards)
is sufficient for many applications, however,
and we assume node dates are not known here.
% This should be obvious, but seems like it needs to be said
In practical settings, we will wish to associate additional
metadata with nodes such as sample identifiers or quality-control metrics.
It is therefore best to think of the
integers used here in the definition of a node as an \emph{identifier},
with which arbitrary additional information can be associated.
% Note: I don't see why we have to apologise for defining nodes as
% integers, that's just the definition. I don't think anyone really
% has a problem with this.
% Concretely defining nodes as integers in this way has some
% significant advantages. In particular, it provides a direct link
% between an ARG and its local trees in terms of a well-defined
% combinatorial object (Section~\ref{sec-ARG-and-local-trees}),
% as well as enabling straightforward connection between
% definitions and practical implementation (Section~\ref{sec-efficiency}).

@@

There is also a certain clarity gained by explicitly modelling nodes
in the inheritance graph as genomes.
Outside of the context of a
% The point being, events are perfectly well defined in the models
% but not in reality
mathematical model, an ``event'' is a slippery concept.
For example, \emph{which} genome along a chain of cell divisions should be
regarded as the one where an event occurred,
or whether multiple coalescences 
within a single individual should be regarded as one or multiple events are
debatable points (Appendix~\ref{sec-cell-lineages-and-args}).
% If we want to have an ARG software ecosystem then such fireside
% discussions won't help.
From the perspective of a concrete data structure,
ideally forming the basis of an ecosystem of interoperable
inference and analysis methods, such debates
are unproductive.

@@

Efficiently propagating and resolving ancestral material for
a sample through a pre-existing graph is a well-studied problem,
and central to recent advances in individual-based forward-time
simulations~\citep{kelleher2018efficient,haller2018tree}.
In contrast to the usual ``retrospective'' view of ARGs
discussed so far, these methods record an ARG forwards in
time in a ``prospective'' manner. Genetic inheritance relationships
and mutations are recorded exhaustively, generation-by-generation,
leading to a rapid build-up of information, much of which
will not be relevant to the genetic ancestry of the current population.
This redundancy is periodically removed using the ``simplify''
algorithm~\citep{kelleher2018efficient}, which propagates and
resolves ancestral material.
Simplification allows efficient subsetting of ARGs (removing both samples and
certain levels of detail about recombination, as discussed in
Section~\ref{sec-ARG-simplification}), and is also the key enabling factor when
incorporating prospective ARGs into
forward-time simulation.
% Define "sample-resolved" here for later use
We refer to a gARG that has been simplified with respect to a set of
samples, such that the inheritance annotations on its edges contain
no non-ancestral material,  as sample-resolved.

@@

\section{ARGs and local trees}
\label{sec-ARG-and-local-trees}
% (We use ``local'' rather than
% ``marginal''
% % could list loads here, but not much point.
% trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping,kelleher2016efficient}
% to avoid the potential confusion with the
% statistical concept of marginal distributions.)
The relationship between an ARG and its corresponding
local trees is subtle and important.
A fundamental property of genetics is that a
given DNA nucleotide is inherited from exactly one parent genome,
both at an organismal and cell-by-cell level
(Appendix~\ref{sec-cell-lineages-and-args}).
These paths of single-parent inheritance give rise,
by definition, to a tree structure.
As a result of recombination, adjacent nucleotides can have
different paths of inheritance, and
an ARG encodes the entire ensemble of local trees along the
% Said "sample of genomes" originally, but using term "genome" in two
% different ways in the same sentence then.
genome for a given set of sample nodes.
% This is for Yan - to me this is a given, but maybe some people feel
% the following is overly formal?
Precisely defining the process by which local trees are extracted
from an ARG is essential to our understanding of how ARGs and local trees
are related, and we require a concrete mathematical structure
to describe the local trees. It is important to note that 
although the following discussion is phrased in terms of the 
gARG encoding,  the arguments apply equally to eARGs
because any eARG can be converted to a 
% Note, being sample-resolve is not necessary for following argument so omitting
gARG without loss of information (Section~\ref{sec-ancestral-material}).

@@

Inferred ARGs are based on the \citet{kreitman1983nucleotide} dataset,
a standard benchmark in the classical ARG
literature.
It consists of 43 biallelic SNPs spanning 2.4Kb of the \emph{D.~melanogaster} ADH
locus on chromosome 2L. Where required for inference purposes we assume
mutation and recombination rates of $5.49\times10^{-9}$
and $2.40463\times10^{-9}$ per site per generation
\citep{schrider2013rates,comeron2012many}
and a constant effective population size of $1720600$ \citep{li2006inferring}, all
values being taken from the \texttt{stdpopsim} library
\citep{adrion2020community,lauterbur2023expanding}.
Software versions were \kwarg\~v1.0, \argweaverD\~(2019),
\tsinfer\~v0.3.1, and \relate\~v1.1.9: full details and replicable code for generating
these figures are available on GitHub (see Data Availability).

@@

Interest in ARG inference methods and downstream applications is
burgeoning, with exciting developments arriving at ever-increasing pace.
Without agreement on basic terminology and some standardisation
on data formats, however, the ARG revolution may falter.
For ARG-based methods to achieve mainstream status, we require
a rich supporting software ecosystem.
Ideally, this would comprise a wide range of 
inference methods specialised to different organisms, types
and scales of data and
% I'm thinking about things like Espalier specifically wanting
% to characterise the recombs, vs Relate mainly wanting to
% get good local trees
targeting particular inference goals, sharing a common ARG interchange format.
The output of these diverse inference methods could then be
processed by numerous downstream applications,
enabling users to freely compare the outputs of different
inference methods on their data, and the result of different
analyses on these inferences.
Earlier efforts to standardise ARG interchange shared this vision,
but did not succeed~\citep{cardona2008extended,mcgill2013graphml}.
Without a well-defined and efficient shared format
for ARG data interchange,
users will suffer from problems with converting data from one program's
output to another, an unfortunate hallmark of population genetics
software for many years~\citep{excoffier2006computer}.
Current methods tend to tightly couple downstream analysis
of the inferred ARG with the inference
itself within the same software package,
which is ultimately not compatible with the widespread use
of ARGs for routine data analysis, and a healthy and diverse software ecosystem.
The gARG encoding described here is a significant generalisation
of classical concepts, capable of describing even the bewildering complexity
of contemporary datasets and capturing important forms of temporal uncertainty.
The \texttt{tskit} library is a high-quality open-source gARG implementation,
with proven efficiency and
scalability~\citep[e.g.][]{anderson2022genes,zhan2023towards},
that is already in widespread use.
Adopting it as a community standard,
and utilising its rich suite of operations,
may accelerate progress and help to
establish ARG methods within mainstream genetics.
